---
title: "Environmental drivers of Margalefidinium IFCB bloom" # title of the notebook
author: "Virginie Sonnet and Jessica Carney" 
date: 2023-09-15
description: 
    This script investigates the environmental forcings of the Margalefidinium bloom in the IFCB in 2018. 
html: 
theme: sandstone
#mainfont: "LM Roman"
fontsize: 0.99em
toc: true # table of contents
toc-depth: 5 
toc-location: left # table of contents on the left 
lightbox: TRUE # allows to click on an image to zoom in the HTML document
embed-resources: true # avoid having dependencies in an extra folder
smooth-scroll: true
editor: visual
code-overflow: wrap
code-fold: false
execute:
    eval: true # run the code chunks FALSE/TRUE
    echo: true # print the source code FALSE/TRUE
    error: true # print error messages FALSE/TRUE
    message: false # print any message FALSE/TRUE
    warning: false # print warning message FALSE/TRUE
    cache: false 
---

```{r}
#| eval: false
#| echo: false

Other outputs: pdf_document, html_notebook, word_document 
The HTML below just says to use the "LM Roman" family as font (kind of Arial?) and if you want the Latex font, you just need "Latin Modern Roman". The line below indicates that the text has to be justified. 
```

```{r}
#| include: false

# make sure that no object is in the environment and collect the garbage 
rm(list=ls())
gc()
```

```{r}
#| results: hide 

# general 
library(tidyverse) 
library(lubridate)

# visualization 
#library(ggpubr)
#library(scales)
library(castr)
library(directlabels)
library(cowplot)
library(wesanderson)
```

# Margalefidinium data

------------------------------------------------------------------------

```{r}
marga <- read_csv("../data/processed_Margalefidinium_conc+vol.csv") %>% 
    # only keep the study time 
    filter(status=="study") %>% 
    # add the bloom period
    mutate(status=case_when(date < ymd("2018-08-11")~"before",
                          date >= ymd("2018-08-11") & date < ymd("2018-08-20")~"pre",
                          date >= ymd("2018-08-20") & date < ymd("2018-08-24")~"during",
                          date >= ymd("2018-08-24")~"after")) %>% 
    # round per hour 
    mutate(date=round_date(date,"hour")) %>% 
    group_by(status,date) %>% 
    summarize(conc=mean(conc,na.rm=TRUE),
              ajvol=mean(adjvol,na.rm=TRUE)) %>% 
    arrange(date)
```

# Environmental time series

------------------------------------------------------------------------

We have collected and processed environmental data from the Bay from different sources and averaged it over the hour to consistently match the IFCB.

```{r}
# environmental data 
envt <- read_csv("../data/NBay_hourly_GSO_nuts.csv",guess_max=180000) %>% 
  filter(date >= ymd("2018-08-01") & date < ymd("2018-09-19")) %>% 
  mutate(status=case_when(date < ymd("2018-08-11")~"before",
                          date >= ymd("2018-08-11") & date < ymd("2018-08-20")~"pre",
                          date >= ymd("2018-08-20") & date < ymd("2018-08-24")~"during",
                          date >= ymd("2018-08-24")~"after")) 
```

To ease the visualization, we can smooth the time series with a window of 17 hours.

```{r}
senvt <- envt %>% 
  select(status,date,temp:nitrate) %>% 
  pivot_longer(temp:nitrate, names_to="var",values_to="val") %>% 
  arrange(var,date) %>% 
  # do the smoothing for each variable
  group_by(var) %>% 
  mutate(sval=smooth(val,k=17)) %>% 
  select(-val) %>% 
  pivot_wider(names_from="var",values_from="sval")
```

## Nutrients

```{r}
# nutrients time series 
nuts <- envt %>% 
  select(status,date,Nitrogen=di_no,Phosphate=di_po,Silicate=si) %>% 
  mutate(status=factor(status,levels=c("before","pre","during","after"))) %>% 
  drop_na() %>% 
  pivot_longer(Nitrogen:Silicate,names_to="var",values_to="val") %>% 
  ggplot(aes(x=date,y=val,group=var)) + 
  geom_line(color="grey50") + 
  geom_point(aes(color=status),size=2) + 
  geom_vline(xintercept=ymd_hms("2018-08-20 00:00:00"),linetype="dashed",
             color="grey80") + 
  geom_vline(xintercept=ymd_hms("2018-08-24 00:00:00"),linetype="dashed",
             color="grey80") + 
  geom_dl(aes(label = var), method = list(dl.trans(x = x + 0.2,y=y+0.1),cex=0.8,
                                          "last.points"),
          color="grey50") + 
  scale_color_manual(values = c("#440154","#577B9E","#35b779","#C3D922")) +
  theme_classic() +
  theme(legend.position=c(0.075,0.6),
        legend.title= element_blank(),
        legend.background = element_blank(),
        legend.box.background=element_rect(color="grey50"),
        legend.box.margin=margin(0,0,0,0),
        legend.spacing.y = unit(0, "mm"),
        text=element_text(size=15),
        axis.title.y=element_text(face="bold",size=10),
        plot.margin = unit(c(0, 0, 0, 0), "cm")) + 
  labs(x="",y=~ bold("Nutrients " (mM/m^2))) + 
  scale_x_datetime(limits=c(ymd_hms("2018-08-01 00:00:00"),ymd_hms("2018-09-18 00:00:00")))
nuts
```

## Temperature

```{r}
temp <- envt %>% 
  select(status,date,temperature=temp) %>% 
  # add the smoothed values 
  left_join(select(senvt,date,stemperature=temp)) %>% 
  pivot_longer(temperature:stemperature,names_to="var",values_to="val") %>%  
  mutate(status=replace(status,var=="temperature","original")) %>% 
  mutate(status=factor(status,levels=c("before","pre","during","after","original"))) %>% 
  ggplot(aes(x=date,y=val,group=var)) + 
  geom_line(color="grey90") + 
  geom_point(aes(color=status,group=var,size=var)) + 
  geom_vline(xintercept=ymd_hms("2018-08-20 00:00:00"),linetype="dashed",
             color="grey80") + 
  geom_vline(xintercept=ymd_hms("2018-08-24 00:00:00"),linetype="dashed",
             color="grey80") + 
  scale_color_manual(values = c("#440154","#577B9E","#35b779","#C3D922",
                                "grey50")) +
  scale_size_manual(values=c(1,0.1)) + 
  theme_classic() +
  theme(legend.position="none",
        text=element_text(size=15),
        axis.title.y=element_text(face="bold",size=10),
        plot.margin = unit(c(0, 0, 0, 0), "cm")) + 
  labs(x="",y="Temperature (°C)")
temp
```

## Salinity

```{r}
# salinity
sal <- envt %>% 
  select(status,date,salinity=sal) %>% 
  left_join(select(senvt,date,ssalinity=sal)) %>% 
  pivot_longer(salinity:ssalinity,names_to="var",values_to="val") %>%  
  mutate(status=replace(status,var=="salinity","original")) %>% 
  mutate(status=factor(status,levels=c("before","pre","during","after","original"))) %>% 
  ggplot(aes(x=date,y=val,group=var)) + 
  geom_line(color="grey90") + 
  geom_point(aes(color=status,group=var,size=var)) + 
  geom_vline(xintercept=ymd_hms("2018-08-20 00:00:00"),linetype="dashed",
             color="grey80") + 
  geom_vline(xintercept=ymd_hms("2018-08-24 00:00:00"),linetype="dashed",
             color="grey80") + 
  #facet_wrap(.~var,scales="free_y") + 
  scale_color_manual(values = c("#440154","#577B9E","#35b779","#C3D922",
                                "grey50")) +
  scale_size_manual(values=c(0.1,1)) + 
  theme_classic() +
  theme(legend.position="none",
        text=element_text(size=15),
        axis.title.y=element_text(face="bold",size=10),
        plot.margin = unit(c(0, 0, 0, 0), "cm")) + 
  labs(x="",y="Salinity")
sal
```

## River input

```{r}
# river input 
riv <- envt %>% 
  select(status,date,flow) %>% 
  mutate(status=factor(status,levels=c("before","pre","during","after"))) %>%
  ggplot(aes(x=date,y=flow)) + 
  geom_line(color="grey90") + 
  geom_point(aes(color=status)) + 
  geom_vline(xintercept=ymd_hms("2018-08-20 00:00:00"),linetype="dashed",
             color="grey80") + 
  geom_vline(xintercept=ymd_hms("2018-08-24 00:00:00"),linetype="dashed",
             color="grey80") + 
  scale_color_manual(values = c("#440154","#577B9E","#35b779","#C3D922",
                                "grey50")) +
  theme_classic() +
  theme(legend.position="none",
        text=element_text(size=15),
        axis.title.y=element_text(size=10),
        plot.margin = unit(c(0, 0, 0, 0), "cm")) + 
  labs(x="",y=~ bold("River flow " (m^3/s)))
riv
```

## Water level range

```{r}
wrange <- envt %>% 
  select(status,date,range) %>% 
  drop_na() %>% 
  mutate(status=factor(status,levels=c("before","pre","during","after"))) %>%
  ggplot(aes(x=date,y=range)) + 
  geom_line(color="grey90") + 
  geom_point(aes(color=status),size=2) + 
  geom_vline(xintercept=ymd_hms("2018-08-20 00:00:00"),linetype="dashed",
             color="grey80") + 
  geom_vline(xintercept=ymd_hms("2018-08-24 00:00:00"),linetype="dashed",
             color="grey80") + 
  scale_color_manual(values = c("#440154","#577B9E","#35b779","#C3D922",
                                "grey50")) +
  theme_classic() +
  theme(legend.position="none",
        text=element_text(size=15),
        axis.title.y=element_text(face="bold",size=10),
        plot.margin = unit(c(0, 0, 0, 0), "cm")) + 
  labs(x="",y="Tidal range (m)")
wrange
```

## Daily precipitation

```{r}
# daily rainfall data
data_pptn <- read_csv("../data/uscrn.csv") %>% 
  mutate(date=ymd(LST_DATE),
         prec=PRECIPITATION) %>% 
  filter(date >= ymd("2018-08-01") & date < ymd("2018-09-18")) %>% 
  mutate(status=case_when(date < ymd("2018-08-11")~"before",
                          date >= ymd("2018-08-11") & date < ymd("2018-08-20")~"pre",
                          date >= ymd("2018-08-20") & date < ymd("2018-08-24")~"during",
                          date >= ymd("2018-08-24")~"after")) %>% 
  select(status,date,prec) %>% 
  group_by(status,date) %>% 
  summarize(prec=sum(prec,na.rm=TRUE))

# plot
pp <- data_pptn %>% 
  select(status,date,prec) %>% 
  mutate(status=factor(status,levels=c("before","pre","during","after"))) %>%
  ggplot(aes(x=date,y=prec,fill=status)) + 
  geom_col() + 
  geom_vline(xintercept=ymd_hms("2018-08-20 00:00:00"),linetype="dashed",
             color="grey80") + 
  geom_vline(xintercept=ymd_hms("2018-08-24 00:00:00"),linetype="dashed",
             color="grey80") + 
  scale_fill_manual(values = c("#440154","#577B9E","#35b779","#C3D922")) +
  theme_classic() +
  theme(legend.position="none",
        text=element_text(size=15),
        axis.title.y=element_text(face="bold",size=10),
        plot.margin = unit(c(0, 0, 0, 0), "cm")) + 
  labs(x="",y="Daily Precipitation (mm)")
pp
```

## Combine

```{r}
plot_grid(temp,sal,pp,riv,wrange,nuts,align="vh",labels=c("A","B","C","D","E","F"),
    ncol=1,nrow=6,label_x=0.97,label_y=1)
```

```{r}
#| eval: false

ggsave(path = "../figures/", filename="Fig7.tiff", width = 7.8, height = 8, device='tiff', dpi=300, units="in",scale=1.2)
```

## Salinity vs. concentration during bloom

```{r}
marga %>% 
  left_join(envt) %>% 
  filter(date > "2018-08-19" & date < "2018-08-24") %>% 
  ggplot(aes(x=sal,y=conc,color=wlevel)) + 
  geom_point() + 
  scale_color_gradientn(colors=wes_palette("Zissou1",5,type = "continuous")) + 
  theme_classic() +
  theme(legend.position=c(0.95,0.75),
        text=element_text(size=15),
        axis.title=element_text(face="bold",size=12),
        legend.title=element_text(size=12),
        plot.margin=unit(c(0.25,0.5,0,0.25), "cm")) + 
  labs(x="Salinity",y="M. polykrikoides (#images/ml)",color="Water \nlevel (m)") 

ggsave(path = "../figures/", filename="FigS4.tiff", width = 4, height = 3, device='tiff', dpi=300, units="in",scale=1.2)
```

# GAM

## Hourly non-detrended

```{r}
library(MASS)
#library(pscl)
library(mgcv)

# let's prepare the data for the first hourly dataset 
data <- envt %>% 
  ungroup() %>% 
  dplyr::select(date,temp,sal,light,pptn,flow,wlevel,wind) %>% 
  right_join(marga) %>% 
  drop_na(temp) %>% 
  mutate(date=with_tz(date,tzone="America/New_York")) %>% 
  mutate(day=floor_date(date,"day"),
         hour=hour(date)) %>% 
  filter(status != "after")
  

visdat::vis_dat(data)

CP<-data[,c("conc")]
t<-data[,c("temp")]
s<-data[,c("sal")]
sr<-data[,c("light")]
prec<-data[,c("pptn")]
tide<-data[,c("wlevel")]
wspd<-data[,c("wind")]
fl<-data[,c("flow")]
dates = data[,c("day")]
bloom_phase=data[,c("status")]
```

### GAM

```{r}
# all variables
fit_tw1 <- gam(conc~temp+sal+wlevel+light+pptn+flow, data= data,
                        family=tw())
summary(fit_tw1)
AIC(fit_tw1)

# best fit 
fit_tw2 <- gam(conc~temp+sal+wlevel, data= data,
                        family=tw())
summary(fit_tw2)
AIC(fit_tw2)
```

```{r}
plot(data$conc,lwd=0.7,col="grey")
#lines(fitted(fit_zigamma3), col="blue",lwd=2)
lines(fit_tw2$fitted.values, col="orange",lwd=2)
legend("topleft", c("GAM-Tweedie"),col=c("orange"),lwd=2,bty="n")
legend(c(0,534),"\n    Data",col="grey",pch=1,bty="n")
```
